CODECHECK: An open-science initiative for the independent execution of computations underlying research articles

### 2021-01-12 codecheck-paper.tex
update principles and register (consistent styling, understandability)

\textbf{1. Codecheckers record but don't investigate or fix.} The
codechecker follows the instructions provided by the author in running
the code. If any steps are unclear, or if code does not run correctly,
the codechecker records this fact, and then communicates this
information to the author. We believe that the job of the codechecker is
not to fix all these problems, but simply to report them, and await a
fix from the author.

\textbf{2. Communication between humans is key.} Some code may just
work without any interaction (e.g. CODECHECK certificate 2020-013
\cite{cert-2020-013}); however, more often there are hidden
dependencies that need to be found, or revised software
installations. Allowing the codechecker and human to communicate
directly and openly with each other should make this process as
constructive as possible. We feel that routing this conversation
(possibly anonymously) through a publisher would slow things down and
reduce the chances for community building as well as the lessons
learned for both author and codecheckers.

\textbf{3. Credit is given to codecheckers.} The value of contributing a
CODECHECK is comparable to providing a scientific peer review. Individual
variations aside, it may also require a similar amount of time. Therefore,
the codechecker's activity should be publicly recorded and mentioned in the
published paper.
The public record can be realised by publishing the certificate in a 
citable form (i.e., with a DOI; this is done within the community codecheck
process), by listing codecheckers on the journal's website or, ideally, by
publishing the checks alongside peer review activities in public databases.

@@

\textbf{5. Open by default and transitional by disposition.} % or "design"?
By default, unless there are strong
reasons to the contrary (e.g., clinically-sensitive data), all code and
data, both from author and codechecker, will be made freely available, i.e., 
under open licenses, before or when the certificate is published.
Note that openness is not required for the paper itself as to accommodate 
journals to transition to sustainable open access models.
The code and data publication should follow community good practices.
By disposition, a CODECHECK workflow should be designed (a) as a recognition of
shortcomings in asserting reproducibility of submissions and (b)
with an understanding of the
complex situation that lead to this. However, establishing a CODECHECK workflow
should go hand in hand with initiatives and a perspective to improve education
around reproducibility of methods and communication of comprehensive 
computational workflows, with the long term goal for a specific community or
journal's target audience to reach a state where the level of verification 
and trust provided by codecheckers becomes part of peer review and regular
merit of all accepted papers.

@@

We have a curated list of 25 certificates available at
\url{https://codecheck.org.uk/register}. These are a mixture of
reproductions of papers for a variety of reasons and fall into three themes:
(1) classic and current papers from computational neuroscience,
(2) COVID-19 modelling preprints, and
(3) GIScience.  
The first theme was the initial set of papers used to explore the concept
of CODECHECK. The idea was to take very well known historical articles from a 
specific domain, in this case the work area of co-author SJE, possibly 
as as a conversation starter for potential collaborating journals.
The second theme is rather opportunistic. As the COVID-19 pandemic swept
over the whole Earth, we hoped to contribute a small part to science's answer
to the unknown virus. The checks were solicited through community interaction
%Several COVID-19 models have been done during the time of peer review of these papers,
and under our initiative rather than requested from journal, but the certificate
has then been acknowledged/cited in the paper \cite{Davies2020-vj}.
The third theme represents co-author DN's service as Reproducibility
Reviewer at the AGILE conference series.
Independent of CODECHECK, the \emph{Reproducible AGILE} Initiative \cite{reproducible_agile}
established a process for reproducing workflows at the AGILE conference
series \cite{nust_improving_2020}.
While using slightly different terms and infrastructure, e.g., ``reproducibility
reports'' are published on OSF instead of certificates on Zenodo, the 
reproducibility reviews adhere to the codecheck principles.
Furthermore, a small number of checks was completed as part of a peer reviews
for GIScience journals.

Table~\ref{tab:register} lists the certificates completed to date,
along with citations to the original papers.
Out of these areas, we would like to briefly highlight three examples here.

\textbf{TODO: SJE, tidy this paragraph.}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Gigascience -- our first paper.  Only visualisation, not machine
  learning (which would have taken several days of compute time).
\item
  report 9 -- covid model of UK lockdown model -- unlike many claims
  in the popular media at the time, the model was reproducible
  [NOORDEN news piece]. 
\item
  Manuel's certificate -- approached us as he requested a certificate
  as paper was entering peer review.
\end{enumerate}

@@

    2020-001 \cite{cert-2020-001} & Machine learning & Code for benchmarking ML classification tool checked post acceptance of manuscript and before its publication in \textit{Gigascience} \cite{Piccolo2020-lo}. \\
    2020-002  \cite{cert-2020-002} & Neuroscience & Code written for this project checked by second project member as demonstration using paper from 1997 showing unsupervised learning from natural images \cite{Hancock1992-mp}. \\
    2020-003  \cite{cert-2020-003} & Neuroscience &  Code written for this project checked by second project member as demonstration using classic paper on models of associative memory \cite{Hopfield1982-mz}. \\
    2020-004  \cite{cert-2020-004} & Neuroscience & Code written for this project checked by second project member as demonstration using classic paper on cart-pole balancing problem \cite{Barto1983-rg}. \\
    2020-005  \cite{cert-2020-005} & Neuroscience & Check of independent reimplementation of spike-timing-dependent plasticity (STDP) model \cite{larisch_re_2019} conducted as demonstration for this paper. \\ % Python ANNarchy
    2020-006  \cite{cert-2020-006} & Neuroscience & Check of independent reimplementation of a generalized linear integrate-and-fire neural model \cite{detorakis_re_2017} conducted as demonstration for this paper. \\
    2020-007  \cite{cert-2020-007} & Neuroscience & Check of independent reimplementation of analysing spike patterns of neurons \cite{hathway_re_2018} conducted as demonstration for this paper. \\
    2020-008  \cite{cert-2020-008} & COVID-19 & Code for modelling of interventions on COVID-19 cases in the UK checked at preprint stage \cite{davies-preprint-2020} and later published \cite{Davies2020-vj}. \\
    2020-009  \cite{cert-2020-009} & COVID-19 & Code for analysis of effectiveness of measures to reduce transmission of SARS-CoV-2 checked as preprint \cite{kucharski-preprint-2020} and later published \cite{kucharski_effectiveness_2020}. \\ % R, check mentiond in acks, but not linked
    2020-010  \cite{cert-2020-010} & COVID-19 & Code for analysis of non-pharmaceutical interventions (Report 9) checked as a preprint \cite{ferguson_report_2020}. \\ % CovidSim
    2020-011  \cite{cert-2020-011} & COVID-19 & Code for modelling of COVID-19 spread across Europe was provided by authors and checked while paper was in press \cite{flaxman_estimating_2020}. \\
    2020-012  \cite{cert-2020-012} & COVID-19 & Code for modelling of COVID-19 spread across the USA was checked as preprint \cite{unwin_report_2020} and later published \cite{unwin_state-level_2020}. \\
    2020-013  \cite{cert-2020-013} & Neuroscience & Code for analsis of rest-activity patterns in people wihout con-mediated vision was checked as a preprint \cite{Spitschan2020.06.02.129502} after direct contact with the authors. \\ % MATLAB
    2020-014  \cite{cert-2020-014} & Neuroscience & Code for analysis of pertubation patterns of neural activity was checked after publication as part of publisher collaboration \cite{Sadeh2020}. \\ % MATLAB
    2020-015  \cite{cert-2020-015} & Neuroscience & Code for a neural network model for human focal seizures was checked after publication as part of publisher collaboration \cite{Liou2020}. \\ % MATLAB
    2020-016  \cite{cert-2020-016} & GIScience & Code for models demonstrating the Modifiable Aral Unit Problem (MAUP) in spatial data science \cite{Brunsdon2020} was checked during peer review. \\ % R
    2020-017  \cite{cert-2020-017} & GIScience & Code for spatial data handling, analysis, and visualisation using a variety of R packages \cite{Bivand2020} was checked after peer review before publication. \\ % R
    2020-018  \cite{cert-2020-018} & GIScience & AGILE conference reproducibility report using a demonstration data subset with cellular automaton for modeling dynamic phenomena \cite{Hojati2020}. \\ % R
    2020-019  \cite{cert-2020-019} & GIScience & AGILE conference reproducibility report with subsampled dataset for reachability analysis of suburban transportation using shared cars \cite{Illium2020}. \\
    2020-020  \cite{cert-2020-020} & GIScience & AGILE conference reproducibility report using a container for checking in-database windows operators for processing spatio-temporal data \cite{Werner2020}. \\ % Python + PostGIS
    2020-021  \cite{cert-2020-021} & GIScience & AGILE conference reproducibility report checking code for comparing supervised machine learning models for spatial nominal entity recognition \cite{Medad2020}. \\ % Python
    2020-022  \cite{cert-2020-022} & GIScience & AGILE conference reproducibility report checking code for visualising text analysis on intents and concepts from geo-analytic questions \cite{Xu2020}. \\ % Python
    2020-023  \cite{cert-2020-023} & GIScience & AGILE conference reproducibility report on analysis of spatial footprints of geo-tagged extreme weather events from social media \cite{Owuor2020}. \\
    2020-024  \cite{cert-2020-024} & Neuroscience & Code for multi-agent system for concept drift detection in electromyography \cite{vieira_driftage_2020} was checked during peer review. \\ % R
    2020-025  \cite{cert-2020-025} & GIScience & Adaptation and application of Local Indicators for Categorical Data (LICD) to archaeological data \cite{carrer_application_2021} was checked after peer review before publication. \\ % R

@@

  \caption{Register of completed certificates as of October 2020.  An interactive version
  is available at \url{http://codecheck.org.uk/register}. % TODO: SJE to complete...

>>>>

\textbf{1. Codecheckers record but don't investigate or fix.} The
codechecker follows the instructions provided by the author in running
the code. If any steps are unclear, or if code does not run correctly,
the codechecker records this fact, and then communicates this
information to the author. We believe that the job of the codechecker is
not to fix all these problems, but simply to report them, and await a
fix from the author.
The level of documentation that is needed for third parties
to reproduce a workflow is extremely hard to get right. Too often this 
uncertainty leads to researchers giving up and not documenting at all.
The conversation with a codechecker fixes this problem.

\textbf{2. Communication between humans is key.} Some code may just
work without any interaction (e.g.,~CODECHECK certificate \texttt{2020-013}
\cite{cert-2020-013}); however, more often there are hidden
dependencies that need to be found, or revised software
installations. Allowing the codechecker and human to communicate
directly and openly with each other should make this process as
constructive as possible. We feel that routing this conversation
(possibly anonymously) through a publisher would slow things down and
reduce the chances for community building as well as the lessons
learned for both author and codecheckers.

\textbf{3. Credit is given to codecheckers.} The value of contributing a
CODECHECK is comparable to providing a scientific peer review. Individual
variations aside, it may also require a similar amount of time. Therefore,
the codechecker's activity should be publicly recorded and mentioned in the
published paper.
The public record can be realised by publishing the certificate in a 
citable form (i.e.,~with a DOI; this is done within the community codecheck
process), by listing codecheckers on the journal's website or, ideally, by
publishing the checks alongside peer review activities in public databases.

@@

\textbf{5. Open by default and transitional by disposition.} % or "design"?
By default, unless there are strong
reasons to the contrary (e.g.,~clinically-sensitive data), all code and
data, both from author and codechecker, will be made freely available, 
i.e.,~under open licenses, before or when the certificate is published.
Note that openness is not required for the paper itself as to accommodate 
journals to transition to sustainable Open Access models.
The code and data publication should follow community good practices.
By disposition, a CODECHECK workflow should be designed (a) as a recognition of
shortcomings in asserting reproducibility of submissions and (b)
with an understanding of the
complex situation that lead to this. However, establishing a CODECHECK workflow
should go hand in hand with initiatives and a perspective to improve education
around reproducibility of methods and communication of comprehensive 
computational workflows, with the long term goal for a specific community or
journal's target audience to reach a state where the level of verification 
and trust provided by codecheckers becomes part of peer review and regular
merit of all accepted papers.

@@

We have a curated list of 25 certificates available at
\url{https://codecheck.org.uk/register}.
Table~\ref{tab:register} lists the certificates completed to date,
along with citations to the original papers.
These CODECHECKs fall into three themes:
(1) classic and current papers from computational neuroscience,
(2) COVID-19 modelling preprints, and
(3) GIScience.

The first theme was the initial set of papers used to explore the concept
of CODECHECK. 
The idea was to take very well known historical articles from a 
specific domain, in this case the work area of SJE, possibly 
as as a conversation starter for potential collaborating journals.

The second theme is rather opportunistic. As the COVID-19 pandemic swept
over the whole Earth, we hoped to contribute a small part to science's answer
to the unknown virus. The checks were solicited through community interaction
%Several COVID-19 models have been done during the time of peer review of these papers,
and under our initiative rather than requested from journal, but some certificates
have since then been acknowledged in reviewed papers
\cite{Davies2020-vj,kucharski_effectiveness_2020}.

The third theme represents DN's service as Reproducibility
Reviewer at the AGILE conference series.
Independent of CODECHECK, the \emph{Reproducible AGILE} Initiative \cite{reproducible_agile}
established a process for reproducing workflows at the AGILE conference
series \cite{nust_improving_2020}.
While using slightly different terms and infrastructure (``reproducibility
reports'' are published on OSF instead of certificates on Zenodo) the 
AGILE reproducibility reviews adhere to the codecheck principles.
Furthermore, a small number of checks was completed as part of a peer reviews
for GIScience journals.

Out of these areas, we would like to briefly highlight three examples here.
\textbf{TODO SJE: tidy this paragraph.}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Gigascience -- our first paper.  Only visualisation, not machine
  learning (which would have taken several days of compute time).
\item
  report 9 -- covid model of UK lockdown model -- unlike many claims
  in the popular media at the time, the model was reproducible
  [NOORDEN news piece]. 
\item
  Manuel's certificate -- approached us as he requested a certificate
  as paper was entering peer review.
\end{enumerate}

@@

    \texttt{2020-001} \cite{cert-2020-001} & Machine learning & Code for benchmarking ML classification tool checked post acceptance of manuscript and before its publication in \textit{Gigascience} \cite{Piccolo2020-lo}. \\
    \texttt{2020-002}  \cite{cert-2020-002} & Neuroscience & Code written for this project checked by second project member as demonstration using paper from 1997 showing unsupervised learning from natural images \cite{Hancock1992-mp}. \\
    \texttt{2020-003}  \cite{cert-2020-003} & Neuroscience &  Code written for this project checked by second project member as demonstration using classic paper on models of associative memory \cite{Hopfield1982-mz}. \\
    \texttt{2020-004}  \cite{cert-2020-004} & Neuroscience & Code written for this project checked by second project member as demonstration using classic paper on cart-pole balancing problem \cite{Barto1983-rg}. \\
    \texttt{2020-005}  \cite{cert-2020-005} & Neuroscience & Check of independent reimplementation of spike-timing-dependent plasticity (STDP) model \cite{larisch_re_2019} conducted as demonstration for this paper. \\ % Python ANNarchy
    \texttt{2020-006}  \cite{cert-2020-006} & Neuroscience & Check of independent reimplementation of a generalized linear integrate-and-fire neural model \cite{detorakis_re_2017} conducted as demonstration for this paper. \\
    \texttt{2020-007}  \cite{cert-2020-007} & Neuroscience & Check of independent reimplementation of analysing spike patterns of neurons \cite{hathway_re_2018} conducted as demonstration for this paper. \\
    \texttt{2020-008}  \cite{cert-2020-008} & COVID-19 & Code for modelling of interventions on COVID-19 cases in the UK checked at preprint stage \cite{davies-preprint-2020} and later published \cite{Davies2020-vj}. \\
    \texttt{2020-009}  \cite{cert-2020-009} & COVID-19 & Code for analysis of effectiveness of measures to reduce transmission of SARS-CoV-2 checked as preprint \cite{kucharski-preprint-2020} and later published \cite{kucharski_effectiveness_2020}. \\ % R, check mentiond in acks, but not linked
    \texttt{2020-010}  \cite{cert-2020-010} & COVID-19 & Code for analysis of non-pharmaceutical interventions (Report 9) checked as a preprint \cite{ferguson_report_2020}. \\ % CovidSim
    \texttt{2020-011}  \cite{cert-2020-011} & COVID-19 & Code for modelling of COVID-19 spread across Europe was provided by authors and checked while paper was in press \cite{flaxman_estimating_2020}. \\
    \texttt{2020-012}  \cite{cert-2020-012} & COVID-19 & Code for modelling of COVID-19 spread across the USA was checked as preprint \cite{unwin_report_2020} and later published \cite{unwin_state-level_2020}. \\
    \texttt{2020-013}  \cite{cert-2020-013} & Neuroscience & Code for analsis of rest-activity patterns in people wihout con-mediated vision was checked as a preprint \cite{Spitschan2020.06.02.129502} after direct contact with the authors. \\ % MATLAB
    \texttt{2020-014}  \cite{cert-2020-014} & Neuroscience & Code for analysis of pertubation patterns of neural activity was checked after publication as part of publisher collaboration \cite{Sadeh2020}. \\ % MATLAB
    \texttt{2020-015}  \cite{cert-2020-015} & Neuroscience & Code for a neural network model for human focal seizures was checked after publication as part of publisher collaboration \cite{Liou2020}. \\ % MATLAB
    \texttt{2020-016}  \cite{cert-2020-016} & GIScience & Code for models demonstrating the Modifiable Aral Unit Problem (MAUP) in spatial data science \cite{Brunsdon2020} was checked during peer review. \\ % R
    \texttt{2020-017}  \cite{cert-2020-017} & GIScience & Code for spatial data handling, analysis, and visualisation using a variety of R packages \cite{Bivand2020} was checked after peer review before publication. \\ % R
    \texttt{2020-018}  \cite{cert-2020-018} & GIScience & AGILE conference reproducibility report using a demonstration data subset with cellular automaton for modeling dynamic phenomena \cite{Hojati2020}. \\ % R
    \texttt{2020-019}  \cite{cert-2020-019} & GIScience & AGILE conference reproducibility report with subsampled dataset for reachability analysis of suburban transportation using shared cars \cite{Illium2020}. \\
    \texttt{2020-020}  \cite{cert-2020-020} & GIScience & AGILE conference reproducibility report using a container for checking in-database windows operators for processing spatio-temporal data \cite{Werner2020}. \\ % Python + PostGIS
    \texttt{2020-021}  \cite{cert-2020-021} & GIScience & AGILE conference reproducibility report checking code for comparing supervised machine learning models for spatial nominal entity recognition \cite{Medad2020}. \\ % Python
    \texttt{2020-022}  \cite{cert-2020-022} & GIScience & AGILE conference reproducibility report checking code for visualising text analysis on intents and concepts from geo-analytic questions \cite{Xu2020}. \\ % Python
    \texttt{2020-023}  \cite{cert-2020-023} & GIScience & AGILE conference reproducibility report on analysis of spatial footprints of geo-tagged extreme weather events from social media \cite{Owuor2020}. \\
    \texttt{2020-024}  \cite{cert-2020-024} & Neuroscience & Code for multi-agent system for concept drift detection in electromyography \cite{vieira_driftage_2020} was checked during peer review. \\ % R
    \texttt{2020-025}  \cite{cert-2020-025} & GIScience & Adaptation and application of Local Indicators for Categorical Data (LICD) to archaeological data \cite{carrer_application_2021} was checked after peer review before publication. \\ % R

@@

  \caption{Register of completed certificates as of December 2020.  An interactive version
  is available at \url{http://codecheck.org.uk/register}.
